Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-26 paper-amia-summit-2019/manuscript.tex
checked typos before submission of the final manuscript

In this study, we focus on the analysis of email-delivered MI, or e-Coaching, to promote healthy eating among young adults. The e-Coaching dataset is composed of email correspondence between a MI counselor and the young adult patient. Unlike transcribed in-person discourse, email correspondence is not clearly segmented into codable speech acts (i.e., utterances). Thus, the unstructured nature of e-Coaching exchanges poses a unique set of analytic challenges. Segmentation of e-Coaching exchanges into textual fragments that correspond to distinct e-Coach and patient communication behaviors is a significant barrier to qualitative analysis of this type of clinical conversation. Automating this task is a unique and challenging problem due to the following reasons:

@@

\section*{Materials and methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of experimental dataset are provided in Table~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and and annotated using the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each instance consists of the $n$ words or punctuation marks after this position and $n$ words or punctuation marks prior to this position, including the position itself. Each sample is assigned a ``new segment'' or ``same segment'' label based on whether there should be a segment break after the current position or not. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (an example of a sequence with assigned labels is provided in the $2^{nd}$ row of Table~\ref{tab:datastat}). \\

\begin{table}[ht]
\centering
\caption{\textbf{Summary of statistics of experimental dataset.}}
\label{tab:datastat}
 \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Sessions}} & \multirow{2}{*}{\textbf{Instances}} & \multicolumn{2}{|c|}{\textbf{Class Segments}} & \multicolumn{2}{|c|}{\textbf{Tokens}} & \multicolumn{2}{|c|}{\textbf{Emails}} & \multicolumn{2}{|c|}{\textbf{Annotation}} \\\cline{3-10}
   &  & \textbf{new}  & \textbf{same} & \textbf{words} & \textbf{punctuations}  & \textbf{patient} & \textbf{provider} & \textbf{method}  & \textbf{codes} \\ \hline    
 49 & 102,917 & 3,138 & 99,779 & 95,777 & 7,140 & 281 & 330 & MYSCOPE & 115 \\ \hline
 \multicolumn{10}{|c|}{segmentation example: Hi[0] XXX[0] ,[0] its[0] good[0] to[0] hear[0] from[0] you[0] .[1] it[0] sounds[0] like[0]...} \\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings as lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Dense real-valued vector (i.e. embedding) is a form of distributed word representation, in which each word is associated with a vector in a low-dimensional vector space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} For words or punctuation marks, which do not have pre-trained embeddings, we utilized the embeddings of the same dimensionality trained on the experimental dataset. CRF utilized lexical features, POS tags and the preceding label.

@@

Figure~\ref{fig:embedding-dimension-mlp} (left) illustrates the performance of CRNN on the task of e-Coaching email segmentation by varying the number of dimensions in pre-trained and corpus-based GloVe \footnote{https://nlp.stanford.edu/projects/glove/} and word2vec embeddings. We observed that the best performance is achieved with pre-trained 300-dimensional word2vec word vectors, when three types of features are used together. Therefore, we report the results for other deep learning models used in this study when 300-dimensional word2vec embedding vectors are utilized. The input layer of MLP consists of a sum of embeddings of $n$ words or punctuation marks before and the sum of embeddings of $n$ words or punctuation marks after the word or punctuation mark, which is the center of a sliding window of $2n$ words or punctuation marks. Figure~\ref{fig:embedding-dimension-mlp} (right) demonstrates the performance of MLP on e-Coaching email segmentation by varying the size of the sliding window. It can be observed that the best performance of MLP is achieved when the size of the sliding window is 4 (or $n=2$). Therefore, MLP results in the remaining experiments are reported when $n$ is set to 2.

@@

\section*{Discussion}
This study is the first effort to design and evaluate machine learning methods for automated segmentation of e-Coaching sessions. Experimental results indicate that CRNN is the best model among all machine learning methods considered for this study. CRNN achieved 0.989 F1-measure overall and 0.825 F1-measure for detecting ``new segment''. Robust performance of CRNN provides an evidence that deep learning models are capable of detecting the boundaries of patient and provider behaviors in email delivered behavioral interventions. Our experiments also highlight the importance of punctuation and POS features along with word embeddings for all machine learning methods employed this study. Although the domain of this study was intentionally focused, we believe that the proposed methods are not limited to e-Coaching and our conclusions can be generalized to other domains, which require discourse segmentation.

Punctuation marks and POS features resulted in significant improvement in the performance of traditional machine learning and deep learning methods. Punctuation features had stronger individual impact on model performance than POS features. In all cases, CRF and MLP performed better, when word embeddings were used in conjunction with punctuation embeddings and POS features. Punctuation embeddings improved the performance of BRNN and CRNN measured by precision, recall and F1-measure, while POS features lowered their AUPR. %We believe that BRNN and CRNN performed poorly with POS features, since POS taggers already utilize previous and following word as features. Since neighboring words are also considered by bi-directional RNN, such redundancy did not contribute to better performance. %We observed that MLP achieved the highest precision which may be related to the fact that MLP poorly learned ``new segment'' and misclassified new segment words to same segments in 30\%-40\% of the time.

@@

Segmentation is the first step of qualitative analysis of unstructured clinical communications, such as e-Coaching. Although several studies have focused on segmentation problem in biomedical context, they are limited to segmenting clinical text in EHR into sections and sentences. No previous studies considered the task of automated segmentation of clinical communications into groups of MI behaviors in the context of unstructured MI sessions. By comparing the performance of machine learning methods for the task of segmentation of e-Coaching emails, we found out that convolutional recurrent neural networks demonstrate the best performance in terms of all performance metrics. Manual segmentation of e-Coaching sessions is a very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify textual segments corresponding to MI behaviors in unstructured clinical dialog, which can then be annotated with an MI behavior annotation model in a pipeline setting. %Automated segmentation and annotation of e-Coaching emails can significantly decrease the time to identify effective communication strategies in e-mail based motivational interviewing.
As future work, we are planning to evaluate the methods presented in this work on the datasets from other behavioral interventions.

>>>>

In this study, we focus on the analysis of email-delivered MI, or e-Coaching, to promote healthy eating among young adults. The e-Coaching dataset is composed of email correspondence between an MI counselor and the young adult patient. Unlike transcribed in-person discourse, email correspondence is not clearly segmented into codable speech acts (i.e., utterances). Thus, the unstructured nature of e-Coaching exchanges poses a unique set of analytic challenges. Segmentation of e-Coaching exchanges into textual fragments that correspond to distinct e-Coach and patient communication behaviors is a significant barrier to qualitative analysis of this type of clinical conversation. Automating this task is a unique and challenging problem due to the following reasons:

@@

\section*{Materials and methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of the experimental dataset are provided in Table~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and annotated using the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each instance consists of the $n$ words or punctuation marks after this position and $n$ words or punctuation marks prior to this position, including the position itself. Each sample is assigned a ``new segment'' or ``same segment'' label based on whether there should be a segment break after the current position or not. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (an example of a sequence with assigned labels is provided in the $2^{nd}$ row of Table~\ref{tab:datastat}). \\

\begin{table}[ht]
\centering
\caption{\textbf{Summary of statistics of the experimental dataset}}
\label{tab:datastat}
 \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Sessions}} & \multirow{2}{*}{\textbf{Instances}} & \multicolumn{2}{|c|}{\textbf{Class Segments}} & \multicolumn{2}{|c|}{\textbf{Tokens}} & \multicolumn{2}{|c|}{\textbf{Emails}} & \multicolumn{2}{|c|}{\textbf{Annotation}} \\\cline{3-10}
   &  & \textbf{new}  & \textbf{same} & \textbf{words} & \textbf{punctuations}  & \textbf{patient} & \textbf{provider} & \textbf{method}  & \textbf{codes} \\ \hline    
 49 & 102,917 & 3,138 & 99,779 & 95,777 & 7,140 & 281 & 330 & MYSCOPE & 115 \\ \hline
 \multicolumn{10}{|c|}{segmentation example: Hi[0] XXX[0] ,[0] its[0] good[0] to[0] hear[0] from[0] you[0] .[1] it[0] sounds[0] like[0]...} \\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings as lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Dense real-valued vector (i.e. embedding) is a form of distributed word representation, in which each word is associated with a vector in a low-dimensional vector space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}\url{https://code.google.com/p/word2vec/}} For words or punctuation marks, which do not have pre-trained embeddings, we utilized the embeddings of the same dimensionality trained on the experimental dataset. CRF utilized lexical features, POS tags and the preceding label.

@@

Figure~\ref{fig:embedding-dimension-mlp} (left) illustrates the performance of CRNN on the task of e-Coaching email segmentation by varying the number of dimensions in pre-trained and corpus-based GloVe \footnote{\url{https://nlp.stanford.edu/projects/glove/}} and word2vec embeddings. We observed that the best performance is achieved with pre-trained 300-dimensional word2vec word vectors, when three types of features are used together. Therefore, we report the results for other deep learning models used in this study when 300-dimensional word2vec embedding vectors are utilized. The input layer of MLP consists of a sum of embeddings of $n$ words or punctuation marks before and the sum of embeddings of $n$ words or punctuation marks after the word or punctuation mark, which is the center of a sliding window of $2n$ words or punctuation marks. Figure~\ref{fig:embedding-dimension-mlp} (right) demonstrates the performance of MLP on e-Coaching email segmentation by varying the size of the sliding window. It can be observed that the best performance of MLP is achieved when the size of the sliding window is 4 (or $n=2$). Therefore, MLP results in the remaining experiments are reported when $n$ is set to 2.

@@

\section*{Discussion}
This study is the first effort to design and evaluate machine learning methods for automated segmentation of e-Coaching sessions. Experimental results indicate that CRNN is the best model among all machine learning methods considered for this study. CRNN achieved 0.989 F1-measure overall and 0.825 F1-measure for detecting ``new segment''. The robust performance of CRNN provides an evidence that deep learning models are capable of detecting the boundaries of patient and provider behaviors in email delivered behavioral interventions. Our experiments also highlight the importance of punctuation and POS features along with word embeddings for all machine learning methods employed this study. Although the domain of this study was intentionally focused, we believe that the proposed methods are not limited to e-Coaching and our conclusions can be generalized to other domains, which require discourse segmentation.

Punctuation marks and POS features resulted in significant improvement in the performance of traditional machine learning and deep learning methods. Punctuation features had a stronger individual impact on model performance than POS features. In all cases, CRF and MLP performed better, when word embeddings were used in conjunction with punctuation embeddings and POS features. Punctuation embeddings improved the performance of BRNN and CRNN measured by precision, recall and F1-measure, while POS features lowered their AUPR. %We believe that BRNN and CRNN performed poorly with POS features, since POS taggers already utilize previous and following word as features. Since neighboring words are also considered by bi-directional RNN, such redundancy did not contribute to better performance. %We observed that MLP achieved the highest precision which may be related to the fact that MLP poorly learned ``new segment'' and misclassified new segment words to same segments in 30\%-40\% of the time.

@@

Segmentation is the first step of qualitative analysis of unstructured clinical communications, such as e-Coaching. Although several studies have focused on the segmentation problem in biomedical context, they are limited to segmenting clinical text in EHR into sections and sentences. No previous studies considered the task of automated segmentation of clinical communications into groups of MI behaviors in the context of unstructured MI sessions. By comparing the performance of machine learning methods for the task of segmentation of e-Coaching emails, we found out that convolutional recurrent neural networks demonstrate the best performance in terms of all performance metrics. Manual segmentation of e-Coaching sessions is a very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify textual segments corresponding to MI behaviors in unstructured clinical dialog, which can then be annotated with an MI behavior annotation model in a pipeline setting. %Automated segmentation and annotation of e-Coaching emails can significantly decrease the time to identify effective communication strategies in e-mail based motivational interviewing.
As future work, we are planning to evaluate the methods presented in this work on the datasets from other behavioral interventions.
